%!TEX root = forallxcam.tex
\thispagestyle{empty}
\phantom{{P.D.\ Magnus}\\
	\emph{University at Albany, State University of New York}\\
	\\
	Modified for the Cambridge Course by:\\Tim Button\\
	\emph{University of Cambridge}}

\vfill

\noindent\hfill {\HUGE\forallxcam}

\vfill


\noindent {P.D.\ Magnus}\\
\emph{University at Albany, State University of New York}\\
\\
Modified for the Cambridge Course by:\\Tim Button\\
\emph{University of Cambridge}




\newpage
\thispagestyle{empty}%
\noindent P.D.\ Magnus would like to thank the people who made this project possible. Notable among these are Cristyn Magnus, who read many early drafts; Aaron Schiller, who was an early adopter and provided considerable, helpful feedback; {and} Bin Kang, Craig Erb, Nathan Carter, Wes McMichael, Selva Samuel,  Dave Krueger, Brandon Lee, and the students of Introduction to Logic, who detected various errors in previous versions of the book.

\
\\
Tim Button would like to thank P.D.\ Magnus for his extraordinary act of generosity, in making \forallx\ available to everyone. Thanks also to Alfredo Manfredini B\"{o}hm, Sam Brain, Felicity Davies, Emily Dyson, Phoebe Hill, Richard Jennings, Justin Niven, Igor Stojanovic and Rob Trueman for noticing mistakes in earlier versions; and to Rob, again, for discussions on how to label gaps hygienically (see \S\ref{s:MultipleGenerality}).

\vfill
\noindent{\copyright\ \ifthenelse{\year=2005}{\number\year}{2005--\number\year} by P.D.\ Magnus and Tim Button. Some rights reserved. 

\
\\This book is based upon P.D.\ Magnus's \forallx, available at \url{fecundity.com/logic}, which was released under a Creative Commons license (CC BY 4.0; full details are available at \url{creativecommons.org/licenses/by/4.0/}). The following is a human-readable summary of (and not a substitute for) that license:

\
\\
{\footnotesize \textbf{License details.} You are free to:\\
	--- \emph{Share}: copy and redistribute the material in any medium or format\\
	--- \emph{Adapt}: remix, transform, and build upon the material
	for any purpose, even commercially.
	\\Under the following terms:\\
	--- \emph{Attribution}: You must give appropriate credit, provide a link to the license, and indicate if changes were made. You may do so in any reasonable manner, but not in any way that suggests the licensor endorses you or your use.}

\
\\In accordance with this license, Tim Button has made changes to P.D.\ Magnus's original text, and added new material, and he offers \forallxcam\ under the same CC BY 4.0 license. This copy of \forallxcam\ was released on \today. The most recent version is available at \url{nottub.com}. 

\
\\
Typeset in \XeLaTeX, using Libertinus Serif and Libertinus Sans. The style for typesetting proofs is based on fitch.sty (v0.4) by Peter Selinger, University of Ottawa.