%!TEX root = forallxsol.tex
\thispagestyle{empty}
\phantom{{P.D.\ Magnus}\\
	\emph{University at Albany, State University of New York}\\
	\\
	Modified for the Cambridge Course by:\\Tim Button\\
	\emph{University of Cambridge}}

\vfill

\phantom{\large\texttt{Solutions Booklet}}

\

\noindent\hfill {\HUGE\forallxcam}

\

\noindent\hfill{\large\texttt{Solutions Booklet}}


\vfill


\noindent {P.D.\ Magnus}\\
\emph{University at Albany, State University of New York}\\
\\
{Tim Button}\\
\emph{University of Cambridge}




\newpage
\thispagestyle{empty}%
\noindent This booklet contains model answers to the practice exercises found in \forallxcam. For several of the questions, there are multiple correct possible answers; in each case, this booklet contains at most one answer.  Answers are given in blue; please contact Tim Button at \url{www.nottub.com} if you have accessibility requirements.

\vfill
\noindent{\copyright\ \ifthenelse{\year=2005}{\number\year}{2005--\number\year} by P.D.\ Magnus and Tim Button. Some rights reserved. 


\
\\This is a solutions booklet for \forallxcam, which is based upon P.D.\ Magnus's \forallx, available at \url{fecundity.com/logic}, and which was released under a Creative Commons license (CC BY 4.0; full details are available at \url{creativecommons.org/licenses/by/4.0/}). The following is a human-readable summary of (and not a substitute for) that license:

\
\\
{\footnotesize \textbf{License details.} You are free to:\\
	--- \emph{Share}: copy and redistribute the material in any medium or format\\
	--- \emph{Adapt}: remix, transform, and build upon the material
	for any purpose, even commercially.
	\\Under the following terms:\\
	--- \emph{Attribution}: You must give appropriate credit, provide a link to the license, and indicate if changes were made. You may do so in any reasonable manner, but not in any way that suggests the licensor endorses you or your use.}

\
\\In accordance with this license, Tim Button has made changes to P.D.\ Magnus's original text, and added new material, and he offers this solutions booklet under the same CC BY 4.0 license. This solutions booklet was released on \today. The most recent version of \forallxcam\ is available at \url{nottub.com}. 

\
\\
Typeset in \XeLaTeX, using Libertinus Serif and Libertinus Sans. The style for typesetting proofs is based on fitch.sty (v0.4) by Peter Selinger, University of Ottawa.